SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source Term Evaluation

### 2018-03-10 paper.tex
edits

As the importance of detailed chemical kinetics for predictive reactive-flow simulations has become recognized~\cite{LU2009192}, the size and complexity of chemical kinetic models describing current and next-generation fuels relevant to transportation and power generation have increased greatly; e.g., a recent biodiesel model~\cite{WESTBROOK2011742} consists of \textasciitilde\num{3500} chemical species, and over \num{17000} reactions.
Moreover, the cost of evaluating the chemical source terms scales linearly with the size of the model, while evaluation and factorization of the chemical kinetic Jacobian---if naively implemented---scale quadratically and cubically with the number of species in the model~\cite{LU2009192}.
These factors often make the use of detailed chemical kinetics prohibitive in practice; e.g., in a recent direct numerical simulation~\cite{Spafford:2010aa} utilizing a 22-species model, evaluation of reaction rates consumed around half of the total run time.
In addition, chemical kinetic Jacobian evaluation and factorization---required for most common implicit integration techniques utilized to deal with numerical stiffness---is a bottleneck for even modestly-sized chemical models in realistic reactive flow simulations, necessitating use of other cost-reduction strategies~\cite{LU2009192}.

A host of techniques exist to lessen the computational demand of chemical kinetic calculations while maintaining fidelity, falling broadly into a few categories; removal of unimportant species and reactions unimportant species and reactions~\cite{Lu:2006bb,Pepiot-Desjardins:2008,Hiremath:2010jw,Niemeyer:2010bt,Curtis:2015}, lumping of species with similar thermo-chemical properties~\cite{Lu:2007,Ahmed:2007fa,Pepiot:2008kq}, and time-scale methods that reduce chemical stiffness~\cite{Maas:1992ws,Lam:1994ws,Lu:2001ve,Gou:2010}.
The interested reader is referred to the recent review of Tur\'anyi and Tomlin~\cite{turanyi2016analysis} for a comprehensive overview.
In addition to the previously mentioned cost reduction methods, effort has gone into improving the integration algorithms and codes that evaluate the chemical kinetics~\cite{Gou:2010,SCHWER2002270,Niemeyer:2016aa,GAO2015287}.
In particular, a carefully derived analytical formulation of the Jacobian can greatly increase the sparsity of the matrix~\cite{SCHWER2002270} and drop the cost of Jacobian evaluation from scaling quadratically with the number of species in the model---for a naive finite-difference implementation---to a linear dependence~\cite{LU2009192}, while Jacobian factorization can be accelerated via sparse-matrix techniques~\cite{superlu99}.
Use of Single-Instruction, Multiple-Data (SIMD) and the related Single-Instruction, Multiple-Thread (SIMT) processors to accelerate chemical kinetic simulations has also been studied, e.g., as in~\cite{CurtisGPU:2017,Sewerin20151375,Shi:2012aa,Niemeyer:2016aa}.

@@

\begin{figure}[phtb]
  \centering
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMD.pdf}
      \caption{Schematic of SIMD processing.  A single compute unit (e.g., a CPU core) contains a vector unit with $N_v$ processing elements, together called a vector-lane.  The vector unit executes a single instruction concurrently on multiple data.}
      \label{F:SIMD}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMT.pdf}
      \caption{Schematic of SIMT processing. A single compute unit (e.g., a GPU streaming multiprocessor) contains many processing elements and hosts many threads, each with an instruction to execute (I1, I2).  Threads with the same instruction execute concurrently on multiple data while the others must wait (leading to thread divergence).}
      \label{F:SIMT}
  \end{subfigure}
\end{figure}

@@

SIMT processing---the foundation of modern GPUs~\cite{lindholm2008nvidia}---is a related computing paradigm that hosts large groups of threads on a single compute element (a streaming multiprocessor in NVIDIA terminology).
A group of threads---typically \num{32}, known as a warp on NVIDIA GPUs---execute the same SIMD instruction on multiple data concurrently (\cref{F:SIMT}).
If some threads must execute a different instruction---e.g., due to if\slash then branching or predication---they are forced to wait and execute later.
This phenomena, known as thread-divergence, is a key consideration for SIMT-processing and can cause serious performance degradation for complicated algorithms~\cite{CurtisGPU:2017}.

@@

Recognizing the need to accelerate Jacobian evaluation and factorization, a number of recent works have been published on analytical chemical kinetic Jacobian evaluation; although as will be discussed at the end of this section, this work offers several key improvements over past efforts.
The \texttt{TChem} software package~\cite{Safta:2011vn} was one of the first packages developed providing this functionality, but has several limitations including: incompatibilities with modern reaction types---i.e., pressure dependent Arrhenius (or P-Log) and Chebyshev reactions---and lack of thread-safety to enable parallel execution~\cite{Curtis2017:tchem}. 
Youssefi~\cite{Youssefi:2011tm} explored the importance of analytical Jacobian matrices for time-scale analysis techniques as well as their effect on computational efficiency in zero-dimensional homogeneous reactor simulations.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility; although incompatible with many combustion codes, a significant increase in Jacobian sparsity was exhibited.
Additionally, in the same work Bisetti provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without need for storage of the full dense Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at expect of strict correctness), and enable tabulation of temperature dependent properties; a \SI{80}{\percent} speedup over a finite-difference-based Jacobian was reported when used in a multidimensional reactive-flow simulation.
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and found speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species) with a Tesla C2075 GPU.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, \texttt{pyJac}, supporting parallel execution on the CPU and SIMT execution on the GPU; a speedup of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian was found on the CPU.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but it was not validated outside the context of use with an implicit-integration technique; in addition the Jacobian was based on an over-constrained system, and a differential algebraic equation (DAE) solver should be used for strict correctness\slash conservation.

@@

Shi et al.~\cite{Shi:2012aa} later extended their work to develop a stabilized-explicit solver on a Tesla C2050; paired with a traditional implicit integrator on quad-core Intel i7 930 which handled the stiffest computational cells; a \SIrange{11}{46}{$\times$} speedup was achieved over the base implicit CPU integrator during simulation of a three-dimensional premixed diesel engine simulation.

@@

Sewerin and Rigopoulos~\cite{Sewerin20151375} implemented a three-stage\slash fifth-order implicit Runge--Kutta method~\cite{wanner1991solving} on both a high-end (Nvidia Quadro 6000) and consumer-grade (Nvidia Quadro 600) GPUs, as well as a standard CPU (two-core, four-thread Intel i5-520M) and a scientific workstation (eight-core, 16-thread Intel Xeon E5-2687W); the high-end GPU was at best \SI{1.8}{$\times$} slower than the workstation CPU (16 threads), while the consumer level GPU was at best \SI{5.5}{$\times$} slower than the corresponding standard CPU (four threads).

@@

In evaluation of the thermo-chemical properties, speedups between \SIrange{8}{13}{$\times$} were on 16 CPU cores, and \SIrange{20}{40}{$\times$} on the GPU; speedups of \textasciitilde\SI{9}{$\times$} and \textasciitilde\SI{25}{$\times$} were found in solution of a partially premixed methanol flame PDE on 16 CPU cores and the GPU respectively.

@@

In contrast, SIMD-based chemical kinetics evaluation\slash integration have been far less studied.
Linford et al.~\cite{Linford:2011} implemented a three-stage, second-order Rosenbrock integrator for atmospheric chemical kinetics on the CPU, GPU and cell croadband engine (CBE)---a specially designed vector processor---and found speedups regularly exceeding~\SI{25}{$\times$} over a serial CPU implementation.
Kroshko and Spiteri~\cite{kroshko2013efficient} implemented a SIMD-vectorized third-order stiff Rosenbrock integrator for atmospheric chemistry on the CBE finding a speedup of \SI{1.89}{$\times$} (a parallel scaling efficiency of \SI{0.94}{$\percent$}) over a serial version of the same code.
Stone et al.,~\cite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including CPUs, GPUs, and MICs.
SIMD-vectorization improved the integrator performance over an OpenMP baseline (which was vectorized by simple compiler hints (i.e., \texttt{\#pragmas}) by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was only \SIrange{1.4}{1.6}{$\times$} faster than the OpenMP baseline due to thread-divergence concerns.

This work will build upon our previous analytical chemical kinetic Jacobian code, \texttt{pyJac}~\cite{Niemeyer:2016aa}, in order to:
\begin{itemize}
 \item Derive and validate a new Jacobian formulation that greatly increases sparsity
 \item Enable cross-platform SIMD\slash SIMT vectorization for the CPU, GPU and other accelerators
 \item Investigate the performance for a wide range chemical kinetic models, specifically looking at speedups due to SIMD\slash SIMT-vectorization as well as relative to the previous version of \texttt{pyJac}, and finally
 \item Discuss future extensions to this work as well as the most promising directions for SIMD\slash SIMT vectorization in reactive-flow simulations
\end{itemize}
To our knowledge, this is the first effort that enables vectorized chemical-kinetic source-term and Jacobian evaluation for any chemical model on the CPU, GPU and other accelerators.

@@

When storing arrays for a chemical kinetic model, the data-storage layout and vectorization patterns are critical to achieving high-performance code.
\Cref{F:mem} depicts an example data array with K-rows and N-columns and index ($i$, $j$) corresponding to the $i$th row and $j$th column.
For example the concentration of species $j$ for the $i$-th thermo-chemical state would be stored in $[C]_{i, j}$ with $1 \le i \le N_{state}$, the number of thermo-chemical states considered for evaluation, and $1 \le j \le \ns/$, the number of species in the model.
The stored concentrations would then have $K = N_{\text{state}}$ rows, and $N = N_{\text{sp}}$ columns.

In the ``C'' (C row-major) format, the concentrations of all species for a single thermo-chemical condition $i$ would be stored sequentially in memory, i.e., with  $[C]_{1, 1}$ in index \num{1}, $[C]_{1, 2}$ in index \num{2} etc., as shown in \cref{F:row_major}.
Conversely in the ``F'' (Fortran column-major) format, the concentrations of a single species $j$ over all thermo-chemical states are adjacent in memory, corresponding to storing $[C]_{1, 1}$ in index \num{1}, $[C]_{2, 1}$ in index \num{2} and so on, as shown in \cref{F:column_major}.
This ordering---along with the device (CPU, GPU, etc.) and vectorization pattern in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In a \textit{shallow}-SIMD\slash SIMT vectorization (also referred to as ``per-thread'' in previous works~\cite{Stone:2013aa} using GPUs), each SIMD-lane (or SIMT-thread) in a compute-unit evaluates thermo-chemical source terms\slash chemical kinetic Jacobian for a different thermo-chemical state.
If the data is stored in ``F''-order, the SIMD-lanes\slash SIMT-thread accessing $[C]_{1, j}\ldots[C]_{N_v, j}$---the concentration of species $j$ for states $1, 2,\ldots N_v$, where $N_v$ is the size of the vector lane for SIMD or the number of threads in a SIMT warp---will load sequential locations in memory.
The first $(j+1)$th species concentration, $[C]_{1, j+1}$, will be $N_{\text{state}}$ memory locations away; this increases the likelihood of cache-misses on the CPU~\cite{gray2000rules}, but conversely is well suited to coalesced memory accesses on the GPU~\cite{NVIDIA:2018}.

In a \textit{deep}-SIMD\slash SIMT vectorization (also referred to as ``per-block'' in previous GPU works), a compute-unit utilizes its SIMD-lanes\slash SIMT-threads cooperatively to evaluate the thermo-chemical source terms for a single thermo-chemical state; thus SIMD-lanes loading $[C]_{1, j} \ldots [C]_{1, j + N_v}$---the species concentrations for state 1, species $j \ldots j + L$---will access sequential memory locations if the data is stored in ``C''-order.
Further, in the ``C''-ordering the furthest difference between any two species concentrations within the same thermo-chemical state is at most $N_{\text{sp}}$, with $N_{\text{sp}} \ll N_{\text{state}}$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU, but may lead to uncoalesced memory-accesses on the GPU.
In addition, a deep vectorization requires both synchronization between SIMD-lanes\slash SIMT-threads via memory-fences\slash barriers (an expensive operation), and may result in SIMD-waste\slash SIMT thread-divergence, caused by different SIMD-lanes\slash SIMT-threads executing different instructions (e.g., resulting from different if\slash then branches)

@@

Finally,~\cref{F:vector_data} shows a vectorized data-ordering that improves the caching patterns of a shallow, ``C''-ordered SIMD vectorization on the CPU (\cref{F:row_major_split}) and a deep, ``F''-ordered SIMT vectorization on the GPU (\cref{F:column_major_split}).
This is accomplished by splitting the slower-varying axis of the data array---columns for ``C''-ordering, and rows for ``F''-ordering---into chunks of size $N_v$ (the SIMD vector-width, or SIMT warp-size) and laying these data out contiguously in memory.
For example, using the data-ordering in~\cref{F:row_major_split}, the concentration of species $j$ for states $i\ldots i+N_v$, $[C]_{i, j} \ldots [C]_{i + N_v, j}$, are contiguous in memory and are followed by the concentration of species $j + 1$ for the same states, $[C]_{i, j + 1} \ldots [C]_{i + N_v, j + 1}$.
This pattern, similar to OpenCL's native vector data-types---e.g., \texttt{double\num{8}} which treats \num{8} contiguous double precision floating point numbers as a single vector datum---ensures that any SIMD operation occurs on data that is contiguous in memory, greatly improving caching and SIMD-throughput.
Conversely, the data-ordering in~\cref{F:column_major_split} is designed to enable coalesced memory accesses for ``F''-ordered, deep-SIMT vectorization on the GPU.
The effects of these various data-ordering and vectorization patterns on performance will be studied in~\cref{S:results}.

@@

This state vector---inspired by~\cite{SCHWER2002270}---has a number of features that recommend its use.
First, the state vector results in highly sparse chemical kinetic Jacobians as will be detailed in~\cref{S:sparsity}.
Second, mass is explicitly conserved in this formulation owing to the fact that the moles of the final species is calculated from the ideal gas law, and the rate of change of the final species from conservation of mass (see \cref{deriv-s:state,deriv-s:source} of the supplemental material respectively); additionally, system is not over-constrained and does not require use of a more complicated DAE solver (as compared to an regular ODE solver) for integration\todo{hansen citation here}.
Finally, the chemical kinetic Jacobian changes relatively little between the \conp/ and \conv/ forms, making maintaining the code-base much simpler.
Although most current combustion codes do not use species moles as a state variable, conversion to\slash from more common mass\slash mole-fractions and moles is straightforward, and once inside integration of an chemical kinetic initial value problem (IVP), the choice of variables no longer matter.

@@

The temperature source term~\cite{TurnsStephenR2012Aitc} is:
\loadeq{dtemp}
where $H_k$, $U_k$, $C_{p,k}$ and $C_{v, k}$ are the enthalpy, internal energy, constant-pressure and constant-volume specific heat of species $k$ in molar units respectively, while $[C]_{k}$ is the concentration, given by:
\begin{equation}
 [C]_{k} = \frac{n_{k}}{V}
\end{equation}
By differentiating the ideal gas law:
\begin{equation}
 PV = n\mathcal{R}T \;,
\end{equation}
where $\mathcal{R}$ is the ideal gas-constant, the volume and pressure source terms can be found:
\loadeq{dparam}

@@

The molar source terms are more complicated due to the derivatives of the net rate of progress---see~\cref{deriv-s:dropi}---and pressure modification terms---\cref{deriv-s:dci}---of the reactions.
The reader is again encouraged to read the supplemental material covering the exact derivations and forms of the reactions solved by \texttt{pyJac}.
The derivative of the molar source term with respect to temperature is:
\loadeq{dnkdotdT}
while, similarly the derivative with respect to the state parameter is:
\loadeq{dnkdotdE}
and the moles of species $j$:
\loadeq{dnkdotdnj}

\subsection{Code Generation and Testing Infrastructure}
\label{s:unittest}
Code generation is enabled by the python package \texttt{loo.py}~\cite{kloeckner_loopy_2014}, which translates user specified psuedo-code and data to OpenCL\slash C-code.
As the name implies, \texttt{loo.py} generates code based on for-loops, this differs from the previous version of \texttt{pyJac}\cite{pyjac16} which generates static code (i.e., fully unrolled loops, with thermodynamic\slash reaction parameters written directly in code rather than stored in arrays).
This caused some issues with large file sizes, long compilation times and even occasionally breaking the \texttt{gcc} and \texttt{nvcc} compilers (the latter issue necessitated breaking the Jacobian\slash source-term evaluations into separate files)\cite{Niemeyer:2016aa}.
The implications of this change will be studied in~\cref{S:jacobian_results}, where the performance of the new version of~\texttt{pyJac} will be compared to the previous versions.

>>>>

As the importance of detailed chemical kinetics for predictive reactive-flow simulations has become recognized~\cite{LU2009192}, the size and complexity of chemical kinetic models describing current and next-generation fuels relevant to transportation and power generation have increased greatly; e.g., a recent biodiesel model~\cite{WESTBROOK2011742} consists of \textasciitilde\num{3500} chemical species, and over \num{17000} reactions.
Moreover, the cost of evaluating the chemical source terms scales linearly with the size of the model, while evaluation and factorization of the chemical kinetic Jacobian---if naively implemented---scale quadratically and cubically with the number of species in the model~\cite{LU2009192}.
These factors often make the use of detailed chemical kinetics prohibitive in practice; e.g., in a recent direct numerical simulation~\cite{Spafford:2010aa} utilizing a 22-species model, evaluation of reaction rates consumed around half of the total run time.
In addition, chemical kinetic Jacobian evaluation and factorization---utilized by most common implicit integration techniques to deal with numerical stiffness---is a bottleneck for even modestly-sized chemical models in realistic reactive flow simulations, necessitating use of other cost-reduction strategies~\cite{LU2009192}.

A host of techniques have been developed to lessen the computational demand of chemical kinetic calculations while maintaining fidelity, these fall broadly into a few categories: removal of unimportant species and reactions unimportant species and reactions~\cite{Lu:2006bb,Pepiot-Desjardins:2008,Hiremath:2010jw,Niemeyer:2010bt,Curtis:2015}, lumping of species with similar thermo-chemical properties~\cite{Lu:2007,Ahmed:2007fa,Pepiot:2008kq}, and time-scale methods that reduce chemical stiffness~\cite{Maas:1992ws,Lam:1994ws,Lu:2001ve,Gou:2010}.
The interested reader is referred to the recent review of Tur\'anyi and Tomlin~\cite{turanyi2016analysis} for a comprehensive overview.
In addition to the previously mentioned cost reduction methods, effort has gone into improving the integration algorithms and codes that evaluate the chemical kinetics~\cite{Gou:2010,SCHWER2002270,Niemeyer:2016aa,GAO2015287}.
In particular, a carefully derived analytical formulation of the Jacobian can greatly increase the sparsity of the matrix~\cite{SCHWER2002270} and drop the cost of Jacobian evaluation from scaling quadratically with the number of species in the model---for a naive finite-difference implementation---to a linear dependence~\cite{LU2009192}, while Jacobian factorization can be accelerated via sparse-matrix techniques~\cite{superlu99}.
Use of Single-Instruction, Multiple-Data (SIMD) and the related Single-Instruction, Multiple-Thread (SIMT) processors to accelerate chemical kinetic simulations has also been studied, e.g., as in~\cite{CurtisGPU:2017,Sewerin20151375,Shi:2012aa,Niemeyer:2016aa}.

@@

\begin{figure}[htb]
  \centering
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMD.pdf}
      \caption{Schematic of SIMD processing.  A single compute unit (e.g., a CPU core) contains a vector unit with $N_v$ processing elements, together called a vector-lane.  The vector unit executes a single instruction concurrently on multiple data.}
      \label{F:SIMD}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMT.pdf}
      \caption{Schematic of SIMT processing. A single compute unit (e.g., a GPU streaming multiprocessor) contains many processing elements and hosts many threads, each with an instruction to execute (I1, I2).  Threads with the same instruction execute concurrently on multiple data while the others must wait (leading to thread divergence).}
      \label{F:SIMT}
  \end{subfigure}
\end{figure}

@@

SIMT processing---the foundation of modern GPUs~\cite{lindholm2008nvidia}---is a related computing paradigm that hosts large numbers of threads on a single compute element (a streaming multiprocessor in NVIDIA terminology).
A group of threads---typically \num{32}, known as a warp on NVIDIA GPUs---execute the same SIMD instruction on multiple data concurrently (\cref{F:SIMT}).
If some threads must execute a different instruction---e.g., due to if\slash then branching or predication---they are forced to wait and execute later.
This phenomena, known as thread-divergence, is a key consideration for SIMT-processing and can cause serious performance degradation for complicated algorithms~\cite{CurtisGPU:2017}.

@@

Recognizing the need to accelerate Jacobian evaluation and factorization, a number of recent works have been published on analytical chemical kinetic Jacobian evaluation; although as will be discussed at the end of this section, this work offers several key improvements over past efforts.
The \texttt{TChem} software package~\cite{Safta:2011vn} was one of the first packages developed providing this functionality, but has several limitations including: incompatibilities with modern reaction types---i.e., pressure dependent Arrhenius (or P-Log) and Chebyshev reactions---and lack of thread-safety to enable parallel execution~\cite{Curtis2017:tchem}. 
Youssefi~\cite{Youssefi:2011tm} explored the importance of analytical Jacobian matrices for time-scale analysis techniques as well as their effect on computational efficiency in zero-dimensional homogeneous reactor simulations.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility; although incompatible with many combustion codes, a significant increase in Jacobian sparsity was exhibited.
Additionally, in the same work Bisetti provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without need for storage of the full dense Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at expect of strict correctness), and enable tabulation of temperature dependent properties; a \SI{80}{\percent} speedup over a finite-difference-based Jacobian was reported when used in a multidimensional reactive-flow simulation.
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and found speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species) with a Tesla C2075 GPU.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, \texttt{pyJac}, supporting parallel execution on the CPU and SIMT execution on the GPU; a speedup of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian was found on the CPU.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but it was not validated outside the context of use with an implicit-integration technique; in addition the Jacobian was based on an over-constrained system, the effect on strict conservation of mass\slash energy was not studied.

@@

Shi et al.~\cite{Shi:2012aa} later extended their work to develop a stabilized-explicit solver on a Tesla C2050, paired with a traditional implicit integrator on quad-core Intel i7 930 which handled the stiffest computational cells; a \SIrange{11}{46}{$\times$} speedup was achieved over the base implicit CPU integrator during simulation of a three-dimensional premixed diesel engine simulation.

@@

Sewerin and Rigopoulos~\cite{Sewerin20151375} implemented a three-stage\slash fifth-order implicit Runge--Kutta method~\cite{wanner1991solving} on both a high-end (NVIDIA Quadro 6000) and consumer-grade (Nvidia Quadro 600) GPUs, as well as a standard CPU (two-core, four-thread Intel i5-520M) and a scientific workstation (eight-core, 16-thread Intel Xeon E5-2687W); the high-end GPU was at best \SI{1.8}{$\times$} slower than the workstation CPU (16 threads), while the consumer level GPU was at best \SI{5.5}{$\times$} slower than the corresponding standard CPU (four threads).

@@

In evaluation of the thermo-chemical properties, speedups between \SIrange{8}{13}{$\times$} were found (over a serial version of the same code) while running on 16 CPU cores, and \SIrange{20}{40}{$\times$} on the GPU; speedups of \textasciitilde\SI{9}{$\times$} and \textasciitilde\SI{25}{$\times$} were found in solution of a partially premixed methanol flame PDE on 16 CPU cores and the GPU respectively.

@@

In contrast, SIMD-based chemical kinetics evaluation\slash integration have been far less studied.
Linford et al.~\cite{Linford:2011} implemented a three-stage, second-order Rosenbrock integrator for atmospheric chemical kinetics on the CPU, GPU and cell croadband engine (CBE)---a specially designed vector processor---and found speedups regularly exceeding~\SI{25}{$\times$} over a serial CPU implementation.
Kroshko and Spiteri~\cite{kroshko2013efficient} implemented a SIMD-vectorized third-order stiff Rosenbrock integrator for atmospheric chemistry on the CBE and found a speedup of \SI{1.89}{$\times$} (a parallel scaling efficiency of \SI{0.94}{$\percent$}) over a serial version of the same code.
Stone et al.,~\cite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including CPUs, GPUs, and MICs.
SIMD-vectorization improved the integrator performance over an OpenMP baseline (which was vectorized by simple compiler hints (i.e., \texttt{\#pragmas}) by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was only \SIrange{1.4}{1.6}{$\times$} faster than the OpenMP baseline due to thread-divergence concerns.

This work will build upon our previous analytical chemical kinetic Jacobian code, \texttt{pyJac}~\cite{Niemeyer:2016aa}, in order to:
\begin{itemize}
 \item Derive and validate a new Jacobian formulation that greatly increases sparsity
 \item Enable cross-platform SIMD\slash SIMT vectorization for the CPU, GPU and other accelerators
 \item Investigate the performance for a wide range chemical kinetic models, specifically looking at speedups due to SIMD\slash SIMT-vectorization as well as relative to the previous version of \texttt{pyJac}, and finally
 \item Discuss future extensions to this work as well as the most promising directions for SIMD\slash SIMT vectorization in reactive-flow simulations
\end{itemize}
To our knowledge, this is the first effort that enables vectorized chemical-kinetic source-term and Jacobian evaluation for any chemical model on a wide-selection of platforms, i.e., the CPU, GPU and other accelerators.

@@

When storing arrays for a chemical kinetic model, the data-storage layout and vectorization patterns are critical to achieving high-performance code.
\Cref{F:mem} depicts an example data array with K-rows and N-columns, with index ($i$, $j$) corresponding to the $i$th row and $j$th column.
For example the concentration of species $j$ for the $i$th thermo-chemical state would be stored in $[C]_{i, j}$ with $1 \le i \le N_{state}$, the number of thermo-chemical states considered for evaluation, and $1 \le j \le \ns/$, the number of species in the model.
The stored concentrations would then have $K = N_{\text{state}}$ rows, and $N = N_{\text{sp}}$ columns.

In the ``C'' (C row-major) format, the concentrations of all species for a single thermo-chemical condition $i$ would be stored sequentially in memory, i.e., with  $[C]_{1, 1}$ in index \num{1}, $[C]_{1, 2}$ in index \num{2} etc., as shown in \cref{F:row_major}.
Conversely in the ``F'' (Fortran column-major) format, the concentrations of a single species $j$ over all thermo-chemical states are adjacent in memory, corresponding to storing $[C]_{1, 1}$ in index \num{1}, $[C]_{2, 1}$ in index \num{2} and so on, as shown in \cref{F:column_major}.
This ordering---along with the device (CPU, GPU, etc.\@) and vectorization pattern in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In a \textit{shallow}-SIMD\slash SIMT vectorization (also referred to as ``per-thread'' in previous works~\cite{Stone:2013aa} using GPUs), each SIMD-lane (or SIMT-thread) in a compute-unit evaluates the thermo-chemical source terms\slash chemical kinetic Jacobian for a different thermo-chemical state.
If the data is stored in ``F''-order, the SIMD-lanes\slash SIMT-thread accessing $[C]_{1, j}\ldots[C]_{N_v, j}$---the concentration of species $j$ for states $1, 2,\ldots N_v$, where $N_v$ is the SIMD vector-width or the number of threads in a SIMT warp---will load sequential locations in memory.
The first $(j+1)$th species concentration, $[C]_{1, j+1}$, will be $N_{\text{state}}$ memory locations away; this increases the likelihood of cache-misses on the CPU~\cite{gray2000rules}, but conversely is well suited to coalesced memory accesses on the GPU~\cite{NVIDIA:2018}.

In a \textit{deep}-SIMD\slash SIMT vectorization (also referred to as ``per-block'' in previous GPU works), a compute-unit utilizes its SIMD-lanes\slash SIMT-threads cooperatively to evaluate the thermo-chemical source terms for a single thermo-chemical state; thus SIMD-lanes loading $[C]_{1, j} \ldots [C]_{1, j + N_v}$---the species concentrations for state 1, species $j \ldots j + N_v$---will access sequential memory locations if the data is stored in ``C''-order.
Further, in the ``C''-ordering the furthest difference between any two species concentrations within the same thermo-chemical state is at most $N_{\text{sp}}$, with $N_{\text{sp}} \ll N_{\text{state}}$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU, but may lead to uncoalesced memory-accesses on the GPU.
In addition, a deep vectorization requires both synchronization between SIMD-lanes\slash SIMT-threads via memory-fences\slash barriers (an expensive operation), and may result in SIMD-waste\slash SIMT thread-divergence, caused by different SIMD-lanes\slash SIMT-threads executing different instructions (e.g., resulting from different if\slash then branches)

@@

Finally,~\cref{F:vector_data} shows a vectorized data-ordering that improves the caching patterns of a shallow, ``C''-ordered SIMD vectorization on the CPU (\cref{F:row_major_split}) and a deep, ``F''-ordered SIMT vectorization on the GPU (\cref{F:column_major_split}).
This is accomplished by splitting the slower-varying axis of the data array---columns for ``C''-ordering, and rows for ``F''-ordering---into chunks of size $N_v$ (the SIMD vector-width, or SIMT warp-size) and laying these data out contiguously in memory.
For example, using the shallow-vectorized ``C''-ordering pictured in~\cref{F:row_major_split}, the concentration of species $j$ for states $i\ldots i+N_v$, $[C]_{i, j} \ldots [C]_{i + N_v, j}$, are contiguous in memory and are followed by the concentration of species $j + 1$ for the same states, $[C]_{i, j + 1} \ldots [C]_{i + N_v, j + 1}$.
This pattern, similar to OpenCL's native vector data-types---e.g., \texttt{double\num{8}} which treats \num{8} contiguous double precision floating point numbers as a single vector datum---ensures that any SIMD operation occurs on data that is contiguous in memory, greatly improving caching and SIMD-throughput.
Conversely, the data-ordering in~\cref{F:column_major_split} is designed to enable coalesced memory accesses for ``F''-ordered, deep-SIMT vectorization on the GPU.
The effects of these various data-ordering and vectorization patterns on performance will be studied in~\cref{S:results}.

@@

This state vector---inspired by~\cite{SCHWER2002270}---has a number of features that recommend its use.
First, the state vector results in highly sparse chemical kinetic Jacobians as will be detailed in~\cref{S:sparsity}.
Second, mass is explicitly conserved in this formulation owing to the fact that the moles of the final species is calculated from the ideal gas law, and the rate of change of the final species from conservation of mass (see \cref{deriv-s:state,deriv-s:source} of the supplemental material respectively); additionally, system is not over-constrained and does not require use of a more complicated DAE solver (as compared to an ODE integrator) for integration\todo{hansen citation here}.
Finally, the chemical kinetic Jacobian changes relatively little between the \conp/ and \conv/ forms, making maintaining the code-base much simpler.
Although most current combustion codes do not use species moles as a state variable, conversion to\slash from more common mass\slash mole-fractions and moles is straightforward, and once inside integration of an chemical kinetic initial value problem (IVP), the choice of variables no longer matter.

@@

The temperature source term~\cite{TurnsStephenR2012Aitc} is:
\loadeq{dtemp}
where $H_k$, $U_k$, $C_{p,k}$ and $C_{v, k}$ are the enthalpy, internal energy, constant-pressure and constant-volume specific heat of species $k$ in molar units respectively, while $[C]_{k}$ is the concentration, given by:
\begin{equation}
 [C]_{k} = \frac{n_{k}}{V}
\end{equation}
By differentiating the ideal gas law:
\begin{equation}
 PV = n\mathcal{R}T \;,
\end{equation}
where $\mathcal{R}$ is the ideal gas-constant in molar units, the volume and pressure source terms can be found:
\loadeq{dparam}

@@

The molar source terms are more complicated due to the derivatives of the net rate of progress---see~\cref{deriv-s:dropi}---and pressure modification terms---\cref{deriv-s:dci}---of the reactions.
The reader is again encouraged to read the supplemental material covering the exact derivations and equations solved by \texttt{pyJac}.
The derivative of the molar source term with respect to temperature is:
\loadeq{dnkdotdT}
while, similarly the derivative with respect to the state parameter is:
\loadeq{dnkdotdE}
and the moles of species $j$:
\loadeq{dnkdotdnj}

\subsection{Code Generation and Testing Infrastructure}
\label{s:unittest}
Code generation is enabled by the python package \texttt{loo.py}~\cite{kloeckner_loopy_2014}, which translates user specified psuedo-code and data to OpenCL\slash C-code.
As the name implies, \texttt{loo.py} generates code based on for-loops, this differs from the previous version of \texttt{pyJac}~\cite{pyjac16} which generates static code (i.e., fully unrolled loops, with thermodynamic\slash reaction parameters written directly in code rather than stored in arrays).
In our previous work,~\cite{Niemeyer:2016aa} this static code generation caused some issues with large file sizes, long compilation times and even occasionally breaking the \texttt{gcc} and \texttt{nvcc} compilers (the latter issue necessitated breaking the Jacobian\slash source-term evaluations into separate files).
The implications of this change will be studied in~\cref{S:jacobian_results}, where the performance of the new version of~\texttt{pyJac} will be compared to the previous versions.
